Visualizing the Geography of Genetic Variants

### 2016-07-01 paper.Rmd
updated typos

The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry. Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography. By surveying the collection of rare variants that an individual carries one can gain insight into the geographic origins of that individual's recent ancestors. One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically-highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.

Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets. For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs). One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast. While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail. A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.

@@

In addition, novel approaches for data visualization have become more widely available.  In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.

>>>>

The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry. Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography. Surveying the collection of rare variants that an individual carries  can provide insight into the geographic origins of that individual's recent ancestors. One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.

Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets. For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs). One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, or a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast. While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail. A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.

@@

In addition, novel approaches for data visualization have become more widely available.  In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing @bostock2011d3.

### 2016-07-01 paper.tex
updated typos

The geographic distribution of rare genetic variants (minor allele
frequency \(\leq .05\)) can be particularly informative in determining
an individual's ancestry. Rare variants tend to be localized in small
geographic regions because they have arisen recently in human history
and have had little time to spread across populations through migration
events {[}4{]}, {[}5{]}. By surveying the collection of rare variants
that an individual carries one can gain insight into the geographic
origins of that individual's recent ancestors. One of us (J.N.) recently
took advantage of the geographic localization of an individual's rare
variants in helping to reaffirm the historically-highly contested Native
American ancestry of the Kennewick Man, an ancient individual from
Washington State \textbf{???}.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide variants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

@@

In addition, novel approaches for data visualization have become more
widely available. In particular, web-based visualization tools, such as
Data Driven Documents (D3.js), offer powerful approaches to realize
these aims through useful methods for interactivity, advantages of
software development in modern web-browsers, a large open-source
development community, and ease of share-ability {[}20{]}.

>>>>

The geographic distribution of rare genetic variants (minor allele
frequency \(\leq .05\)) can be particularly informative in determining
an individual's ancestry. Rare variants tend to be localized in small
geographic regions because they have arisen recently in human history
and have had little time to spread across populations through migration
events {[}4{]}, {[}5{]}. Surveying the collection of rare variants that
an individual carries can provide insight into the geographic origins of
that individual's recent ancestors. One of us (J.N.) recently took
advantage of the geographic localization of an individual's rare
variants in helping to reaffirm the historically contested Native
American ancestry of the Kennewick Man, an ancient individual from
Washington State \textbf{???}.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide variants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}6{]}, {[}7{]}, or a model-based
clustering method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While
these approaches are extremely valuable, researchers can use them to
often without inspecting the underlying variant data in more detail. A
natural approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

@@

In addition, novel approaches for data visualization have become more
widely available. In particular, web-based visualization tools, such as
Data Driven Documents (D3.js), offer powerful approaches to realize
these aims through useful methods for interactivity, advantages of
software development in modern web-browsers, a large open-source
development community, and ease of sharing {[}20{]}.
